Ship-based lidar evaluation of Southern Ocean low clouds in the storm-resolving general circulation model ICON and the ERA5 and MERRA-2 reanalyses

### 2025-07-21 manuscript_agu_rev1.tex
Fix grammar

\begin{abstract}
Global storm resolving models (GSRMs) represent the next generation of global climate models. One of them is a 5-km Icosahedral Nonhydrostatic Weather and Climate Model (ICON). Its high resolution means that parameterizations of convection and clouds, including subgrid-scale clouds, are omitted, relying on explicit simulation but necessarily utilizing microphysics and turbulence parameterizations. Standard-resolution (10--100~km) models, which use convection and cloud parameterizations, have substantial cloud biases over the Southern Ocean (SO), adversely affecting radiation and sea surface temperature. The SO is dominated by low clouds, which cannot be observed accurately from space due to overlapping clouds, attenuation, and ground clutter. We evaluated SO clouds in ICON and the ERA5 and MERRA-2 reanalyses using approximately 2400 days of lidar observations and 2300 radiosonde profiles from 31 voyages and Macquarie Island station during 2010--2021, compared with the models using a ground-based lidar simulator. We found that ICON and the reanalyses underestimate the total cloud fraction by about 10 and 20\%, respectively. ICON and ERA5 overestimate the cloud occurrence peak at about 500~m, associated with underestimated lower tropospheric stability and overestimated lifting condensation level. The reanalyses strongly underestimate fog and very low-level clouds, and MERRA-2 underestimates cloud occurrence at almost all heights. Outgoing shortwave radiation is overestimated in MERRA-2, implying a ``too few, too bright'' cloud problem. SO cloud and fog biases are a substantial issue in the analyzed models and result in shortwave and longwave radiation biases.
\end{abstract}

@@

Increasing climate model spatial resolution is one way of improving the accuracy of the representation of the climate system in models \cite{mauritsen2022}. It has been practiced since the advent of climate modeling as more computational power, memory, and storage capacity become available. It is, however, often not as easy as changing the grid size because of the complex interplay between model dynamics and physics, which necessitates adjusting and tuning all components together. Increasing resolution is, of course, limited by the available computational power and a trade-off with increasing parameterization complexity, which is another way of improving model accuracy. Current computational availability and acceleration from general-purpose computing on graphics processing units (GPUs) has progressed to enable km-scale (also called k-scale) Earth system models (ESMs) and coupled atmosphere--ocean general circulation models (AOGCMs) for research today and will become operational in the future. Therefore, it represents a natural advance in climate modeling. Global storm-resolving models (GSRMs) are emerging as a new front in the development of high-resolution global climate models, with horizontal grid resolutions of about 2--8~km \cite{satoh2019,stevens2019}. This resolution is enough to resolve mesoscale convective storms, but smaller-scale convective plumes and cloud structure remain unresolved. At an approximately 5-km scale, non-hydrostatic processes also become important \cite{weisman1997}, and for this reason such models are generally non-hydrostatic. The terms global cloud-resolving models or global convection-permitting/-resolving models are also sometimes used interchangeably with GSRMs but imply that clouds or convection are resolved explicitly, which is not entirely true for GSRMs, as this would require an even higher horizontal resolution \cite{satoh2019}. Representative of these efforts is the DYnamics of the Atmospheric general circulation Modeled On Non-hydrostatic Domains (DYAMOND) project \cite{stevens2019,dyamond}, which is an intercomparison of nine global GSRMs over two 40-day time periods in summer (1~August--10~September 2016) and winter (20~January--1~March 2020). A new one-year GSRM intercomparison is currently proposed by \citeA{takasuka2024}, with the hope of also evaluating the seasonal cycle and large-scale circulation. An alternative to using a computationally costly GSRM is to train an artificial neural network on GSRM output and use it for subgrid-scale clouds, as done with the GSRM ICON by \citeA{grundner2022} and \citeA{grundner2023}.

The main aim of this study is to evaluate the GSRM version of ICON developed by the nextGEMS project \cite{nextgems,segura2025}. ICON is developed and maintained jointly by Deutscher Wetterdienst, the Max-Planck-Institute for Meteorology, Deutsches Klimarechenzentrum (DKRZ), Karlsruhe Institute of Technology, and the Center for Climate Systems Modeling. Our aim is to quantify how well the GSRM ICON simulates clouds over the Southern Ocean (SO), particularly in light of the fact that subgrid-scale clouds and convection are not parameterized in this model. This region is mostly dominated by boundary layer clouds generated by shallow convection, and these are problematic to observe by spaceborne lidars and radars, which are affected by attenuation by overlapping and thick clouds \cite{mace2009,medeiros2010} and ground clutter \cite{marchand2008}, respectively. Specifically, the radar on CloudSat and lidar on the Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation (CALIPSO), neither of which are now operational, are affected by the above-mentioned issues, resulting in a strong underestimation of cloud occurrence below 2 km in a merged CloudSat--CALIPSO product relative to ground-based lidar observations at McMurdo Station \cite{mcerlich2021}. Removing situations with higher overlapping clouds could enable a less biased comparison of low clouds. We hypothesize that this, in turn, can lead to systematic biases in low clouds in climate models, which are frequently evaluated against CloudSat--CALIPSO products. Reanalyses can also suffer from cloud biases, as these are usually parameterized in their atmospheric component and also in regions where input observations are sparse. This makes them a problematic reference for clouds over the SO, and any biases relative to a reanalysis should be interpreted with caution. Instead, we chose to use a large set of ship-based observations conducted with ceilometers and lidars on board the RV \emph{Polarstern} and other voyages and a station as a reference for the model evaluation. Altogether, we analyzed approximately 2400 days of data from 31 voyages and one sub-Antarctic station covering diverse longitudes and latitudes of the SO. To achieve a like-for-like comparison with the models (ICON, MERRA-2, and ERA5), we used a ground-based lidar simulator called the Automatic Lidar and Ceilometer Framework [ALCF; \citeA{kuma2021}]. We contrasted the results with ERA5 \cite{era5} and the Modern-Era Retrospective analysis for Research and Applications, Version~2 [MERRA-2; \citeA{gelaro2017}].

The nextGEMS project focuses on the research and development of GSRMs at multiple modeling centers and universities in Europe. The project also develops GSRM versions of the Icosahedral Nonhydrostatic Weather and Climate Model (ICON; \citeA{hohenegger2023}), the Integrated Forecasting System [IFS; \citeA{ifs2023}], and their ocean components at eddy-resolving resolutions: ICON-O \cite{korn2022} coupled with ICON and Finite-Element/volumE Sea ice-Ocean Model [FESOM; \citeA{wang2014a}] and Nucleus for European modeling of the Ocean [NEMO; \citeA{madec2023}] coupled with IFS. The project has so far produced ICON and IFS simulations with three development versions called Cycle 1--3 and a pre-final version, with a final production version planned by the end of the project. nextGEMS is not the only project developing GSRMs; other GSRMs (or GSRM versions of climate models) currently in development include: Convection-Permitting Simulations With the E3SM Global Atmosphere Model [SCREAM; \citeA{caldwell2021}], Non-hydrostatic Icosahedral Atmospheric Model [NICAM; \citeA{satoh2008}], Unified Model (UM), eXperimental System for High-resolution modeling for Earth-to-Local Domain [X-SHiELD; \citeA{shield}], Action de Recherche Petite Echelle Grande Echelle-NonHydrostatic version [ARPEGE-NH; \citeA{bubnova1995,voldoire2017}], Finite-Volume Dynamical Core on the Cubed Sphere [FV3, \citeA{lin2004}], the National Aeronautics and Space Administration (NASA) Goddard Earth Observing System global atmospheric model version 5 [GEOS5; \citeA{putman2011}], Model for Prediction Across Scales [MPAS; \citeA{skamarock2012}], and System for Atmospheric Modeling [SAM; \citeA{khairoutdinov2003}].

Multiple cloud properties have an effect on shortwave (SW) and longwave (LW) radiation. To first order, the total cloud fraction, cloud phase, and the liquid and ice water path are the most important cloud properties influencing SW and LW radiation. These properties are in turn influenced by the atmospheric thermodynamics, convection and circulation, and both the indirect and direct effects of aerosols. Second-order effects on SW and LW radiation are associated with the cloud droplet size distribution, ice crystal habit, cloud lifetime, and direct radiative interaction with aerosols \cite{boucher2013}. In the 6$^\mathrm{th}$ phase of the Coupled Model Intercomparison Project [CMIP6; \citeA{eyring2016}], the cloud feedback has increased relative to CMIP5 \cite{zelinka2020}, especially in the Southern Hemisphere mid-to-high latitudes, which is one of the main reasons for the higher climate sensitivity of CMIP6 models.

@@

In the following text, we refer to the SO as ocean regions south of 40°S, low-latitude SO as 40--55°S, and high-latitude SO as south of 55°S, all the way to the Antarctic coast. The reason for this dividing latitude is to split the SO into about two equal zones, as well as the results by \citeA{schuddeboom2021} (Fig.~2b) which show a contrast in CMIP model radiation biases. \citeA{schuddeboom2019} (Fig.~2) and \citeA{kuma2020} (Fig.~3) also show contrasting radiation biases in the Hadley Centre Global Environmental Model, which is also supported by \citeA{cesana2022}, displaying contrasting cloud biases due to the 0°C isotherm reaching the surface at 55°S. The findings of \citeA{niu2024}, however, support a different dividing line of 62°S based on cloud condensation nuclei concentration.

SO radiation biases have been relatively large and systematic compared to the rest of the globe since at least CMIP3 \cite{trenberth2010,bodas-salcedo2012}, and the SO SW cloud radiative effect (CRE) bias is still positive in eight CMIP6 models analyzed by \citeA{schuddeboom2021} over the high-latitude SO, whereas over the low-latitude SO it tends to be more neutral or negative in some models. Too much absorbed SW radiation over the SO was also identified in the GSRM SCREAM \cite{caldwell2021}. Compensating biases are possible, such as the ``too few too bright'' cloud bias, characterized by too small a cloud fraction and too large a cloud albedo \cite{wall2017,kuma2020}, previously described by \citeA{webb2001}, \citeA{weare2004}, \citeA{zhang2005}, \citeA{karlsson2008}, \citeA{nam2012}, \citeA{klein2013}, and \citeA{bender2017} in other regions and models, which means that a model can maintain a reasonable SW radiation balance by reflecting too much SW radiation from clouds, but these cover too small an area. A study by \citeA{konsta2022} showed that this type of bias is still present in six analyzed CMIP6 models in tropical marine clouds, using the GCM-Oriented CALIPSO Cloud Product [CALIPSO--GOCCP; \citeA{chepfer2010}] and Polarization \& Anisotropy of Reflectances for Atmospheric Sciences coupled with Observations from a Lidar [PARASOL; \citeA{lier2008}] as a reference. They suggest improper simulation of subgrid-scale cloud heterogeneity as a cause. Compensating cloud biases in the Australian Community Climate and Earth System Simulator (ACCESS) – Atmosphere-only model version 2 (AM2) over the SO were analyzed by \citeA{fiddes2022} and \citeA{fiddes2024}. \citeA{possner2022} showed that over the SO, the DYAMOND GSRM ICON underestimates low-level cloud fraction on the order of 30\% and overestimates net downward TOA SW radiation by approximately 10 Wm$^\mathrm{-2}$ in the highest model resolution run (2.5~km). \citeA{zhao2022} reported a similar SW radiation bias in five analyzed CMIP6 models over the high-latitude SO and an underestimation of the total cloud fraction on the order of 10\% over the entire 40--60°S SO. Recently, \citeA{ramadoss2024} analyzed 48~hours of km-scale ICON limited-area model NWP simulations over an SO region adjacent to Tasmania against the Clouds, Aerosols, Precipitation, Radiation, and atmospherIc Composition Over the southeRn oceaN (CAPRICORN) voyage cloud and precipitation observations \cite{mcfarquhar2021}. They found the ICON cloud optical thickness was underestimated relative to Himawari‐8 satellite observations but also identified large differences in cloud top phase.

In general, sea surface temperature (SST) biases in the SO can originate either in the atmosphere \cite{hyder2018}, caused by too much shortwave heating of the surface or too little longwave cooling of the surface, such as in situations of too much cloud cover or cloud optical thickness, or in the ocean circulation. Interactions of both are also possible; for example, SST affecting clouds and clouds affecting the surface radiation. Using the European Centre for Medium-Range Weather Forecasts (ECMWF) Reanalysis 5 (ERA5) as a reference, \citeA{zhang2023} have shown that SST biases have improved in CMIP6 compared to CMIP5, with SST overall increasing in CMIP6. However, over the SO, this resulted in an even higher positive bias, especially in the Atlantic Ocean (AO) sector of the SO, increasing by up to 1°C. \citeA{luo2023} identified that the SO SST bias in an ensemble of 18 CMIP6 models originates not from the surface heat and radiation fluxes (using reanalyses as a reference) but from a warm bias in the Northern Atlantic Deep Water.

\begin{figure}[b!]
\centering
\includegraphics[width=\textwidth]{img/map_fig.pdf}
\caption{
\textbf{(a)} A map showing the tracks of 31 voyages of RV \emph{Polarstern}, RSV \emph{Aurora Australis}, RV \emph{Tangaroa}, RV \emph{Nathaniel B. Palmer}, and HMNZS \emph{Wellington} and one sub-Antarctic station (Macquarie Island) analyzed here. The tracks cover Antarctic sectors south of South America, the Atlantic Ocean, Africa, Australia, and New Zealand in the years 2010--2021 (inclusive). The dotted and dashed lines at 40°S and 55°S delineate the Southern Ocean area of our analysis and its partitioning into two subsets, respectively. A photo of \textbf{(b)} RV \emph{Polarstern} (© Folke Mehrtens, Alfred-Wegener-Institut), \textbf{(c)} Lufft CHM~15k installed on RV \emph{Tangaroa} (© Peter Kuma, University of Canterbury), \textbf{(d)} Vaisala CL51 (© Jeff Aquilina, Bureau of Meteorology), \textbf{(e)} Vaisala CT25K at Macquarie Island (© Simon P.\ Alexander, Australian Antarctic Division).
}
\label{fig:map}
\end{figure}

@@

Apart from lidar observations, radiosondes were launched on weather balloons at regular synoptic times on the RV \emph{Polarstern}, MARCUS, NBP17024, TAN1702, and TAN1802 campaigns, measuring pressure, temperature, relative humidity (RH), and the global navigation satellite system coordinates. In total, about 2300 radiosonde profiles south of 40°S were available. Spatially and temporally collocated profiles were taken from the models. Because the time period covered by the ICON model output (2021--2024) was different from the time period covered by the observations (2010--2021), when comparing with ICON, we first had to remap the observation time to model time by taking the same time relative to the start of the year. Consequently, we also had four virtual/model profiles (one for each year from 2021 to 2024) for each observed profile. Derived thermodynamic [virtual potential temperature ($\theta_v$), lifting condensation level (LCL), etc.] and dynamic physical quantities (wind speed and direction) for the measured vertical profiles were calculated with the program radiosonde tool [rstool; \citeA{rstool}]. Surface meteorological quantities were measured continuously by an onboard automatic weather station or individual instruments.

@@

An overview of the analyzed campaigns (voyages and stations). Start, end, and the number of days (UTC; inclusive) refer to the time period when the vessel was south of 40°S. Abbreviations: ceilometer (ceil.), Australia (AU), New Zealand (NZ), South America (SA), Atlantic Ocean (AO), and Africa (AF). The number of days is rounded to the nearest integer. CL51/31 indicates CL51 configured to emulate CL31. Missing days in the ceilometer data were HMNZSW16 (7 days): 24--27 November, 10~December, and 16--17~December 2016; MARCUS (3 days): 8, 10~November, and 10~December 2017; MICRE (9~days): 7--8, 29~June, 5, 16~July, 15~August, 17~October 2016, 11~February, and 21~March 2017; and TAN1502 (1~day): 24~January.

@@

A coupled (atmosphere--ocean) GSRM version of the ICON model is in development as part of the nextGEMS project \cite{hohenegger2023}. ICON is a very flexible model, allowing for simulations ranging from coarse-resolution ESM simulations, GSRM simulations, limited area model simulations, and large eddy simulations (LES) for both weather prediction and climate projections. ICON uses the atmospheric component ICON-A \cite{giorgetta2018}, whose physics is derived from ECHAM6 \cite{stevens2013}, and the ocean component ICON-O \cite{korn2022}. Earlier runs of the GSRM ICON from DYAMOND were evaluated by \citeA{mauritsen2022}.

Here, we use a free-running (i.e., the weather conditions in the model do not correspond to reality) coupled GSRM simulation made for the purpose of climate projection. nextGEMS has so far produced four cycles of model runs. We used a Cycle 3 run \emph{ngc3028} produced in 2023 \cite{nextgems2023a,nextgems2023b} for a model time period of 20 January 2020 to 22 July 2025, of which we analyzed the period 2021--2024 (inclusive). The horizontal resolution of ngc3028 is about 5 km. The model output is available on 90 vertical levels and 3-hourly instantaneous temporal resolution.

Unlike current general circulation models (GCMs), the storm-resolving version of ICON does not use convective and cloud parameterization but relies on explicit simulation of convection and clouds on the model grid. Subgrid-scale clouds are not resolved, and the grid cell cloud fraction is always either 0 or 100\%. While this makes the code development simpler without having to rely on uncertain parameterizations, it can miss smaller-scale clouds below the grid resolution. Turbulence and cloud microphysics have to be parameterized in this model as in other models, and aerosols are derived from a climatology. To account for the radiative effects of subgrid-scale clouds, a cloud inhomogeneity factor is introduced in the model, which scales down the cloud liquid water for radiative calculations. It ranges from 0.4 at lower tropospheric stability (LTS) of 0~K to 0.8 at 30~K. In addition, turbulent mixing in the Smagorinsky scheme was adjusted to allow mixing or entrainment in situations of no mixing under the traditional scheme, affecting stratocumulus clouds but not trade wind clouds \cite{segura2025}.

>>>>

\begin{abstract}
Global storm resolving models (GSRMs) represent the next generation of global climate models. One of them is a 5-km Icosahedral Nonhydrostatic Weather and Climate Model (ICON). Its high resolution means that parameterizations of convection and clouds, including subgrid-scale clouds, are omitted, relying on explicit simulation but necessarily utilizing microphysics and turbulence parameterizations. Standard-resolution (10–100~km) models, which use convection and cloud parameterizations, have substantial cloud biases over the Southern Ocean (SO), adversely affecting radiation and sea surface temperature. The SO is dominated by low clouds, which cannot be observed accurately from space due to overlapping clouds, attenuation, and ground clutter. We evaluated SO clouds in ICON and the ERA5 and MERRA-2 reanalyses using approximately 2400 days of lidar observations and 2300 radiosonde profiles from 31 voyages and Macquarie Island station during 2010–2021, compared with the models using a ground-based lidar simulator. We found that ICON and the reanalyses underestimate the total cloud fraction by about 10 and 20\%, respectively. ICON and ERA5 overestimate the cloud occurrence peak at about 500~m, associated with underestimated lower tropospheric stability and overestimated lifting condensation level. The reanalyses strongly underestimate fog and very low-level clouds, and MERRA-2 underestimates cloud occurrence at almost all heights. Outgoing shortwave radiation is overestimated in MERRA-2, implying a ``too few, too bright'' cloud problem. SO cloud and fog biases are a substantial issue in the analyzed models and result in shortwave and longwave radiation biases.
\end{abstract}

@@

Increasing climate model spatial resolution is one way of improving the accuracy of the representation of the climate system in models \cite{mauritsen2022}. It has been practiced since the advent of climate modeling as more computational power, memory, and storage capacity become available. It is, however, often not as easy as changing the grid size because of the complex interplay between model dynamics and physics, which necessitates adjusting and tuning all components together. Increasing resolution is, of course, limited by the available computational power and a trade-off with increasing parameterization complexity, which is another way of improving model accuracy. Current computational availability and acceleration from general-purpose computing on graphics processing units has progressed to enable km-scale (also called k-scale) Earth system models (ESMs) and coupled atmosphere–ocean general circulation models for research today and will become operational in the future. Therefore, it represents a natural advance in climate modeling. Global storm-resolving models (GSRMs) are emerging as a new front in the development of high-resolution global climate models, with horizontal grid resolutions of about 2–8~km \cite{satoh2019,stevens2019}. This resolution is enough to resolve mesoscale convective storms, but smaller-scale convective plumes and cloud structure remain unresolved. At an approximately 5-km scale, non-hydrostatic processes also become important \cite{weisman1997}, and for this reason such models are generally non-hydrostatic. The terms global cloud-resolving models or global convection-permitting/-resolving models are also sometimes used interchangeably with GSRMs but imply that clouds or convection are resolved explicitly, which is not entirely true for GSRMs, as this would require an even higher horizontal resolution \cite{satoh2019}. Representative of these efforts is the DYnamics of the Atmospheric general circulation Modeled On Non-hydrostatic Domains (DYAMOND) project \cite{stevens2019,dyamond}, which is an intercomparison of nine global GSRMs over two 40-day time periods in summer (1~August–10~September 2016) and winter (20~January–1~March 2020). A new one-year GSRM intercomparison is currently proposed by \citeA{takasuka2024}, with the hope of also evaluating the seasonal cycle and large-scale circulation. An alternative to using a computationally costly GSRM is to train an artificial neural network on GSRM output and use it for subgrid-scale clouds, as done with the GSRM ICON by \citeA{grundner2022} and \citeA{grundner2023}.

The main aim of this study is to evaluate the GSRM version of ICON developed by the nextGEMS project \cite{nextgems,segura2025}. ICON is developed and maintained jointly by Deutscher Wetterdienst, the Max-Planck-Institute for Meteorology, Deutsches Klimarechenzentrum (DKRZ), Karlsruhe Institute of Technology, and the Center for Climate Systems Modeling. Our aim is to quantify how well the GSRM ICON simulates clouds over the Southern Ocean (SO), particularly in light of the fact that subgrid-scale clouds and convection are not parameterized in this model. This region is mostly dominated by boundary layer clouds generated by shallow convection, and these are problematic to observe by spaceborne lidars and radars, which are affected by attenuation by overlapping and thick clouds \cite{mace2009,medeiros2010} and ground clutter \cite{marchand2008}, respectively. Specifically, the radar on CloudSat and lidar on the Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation (CALIPSO), neither of which are now operational, are affected by the above-mentioned issues, resulting in a strong underestimation of cloud occurrence below 2 km in a merged CloudSat–CALIPSO product relative to ground-based lidar observations at McMurdo Station \cite{mcerlich2021}. Removing situations with higher overlapping clouds could enable a less biased comparison of low clouds. We hypothesize that this, in turn, can lead to systematic biases in low clouds in climate models, which are frequently evaluated against CloudSat–CALIPSO products. Reanalyses can also suffer from cloud biases, as these are usually parameterized in their atmospheric component and also in regions where input observations are sparse. This makes them a problematic reference for clouds over the SO, and any biases relative to a reanalysis should be interpreted with caution. Instead, we chose to use a large set of ship-based observations conducted with ceilometers and lidars on board the research vessel (RV) \emph{Polarstern} and other voyages and a station as a reference for the model evaluation. Altogether, we analyzed approximately 2400 days of data from 31 voyages and one sub-Antarctic station covering diverse longitudes and latitudes of the SO. To achieve a like-for-like comparison with the models (ICON, MERRA-2, and ERA5), we used a ground-based lidar simulator called the Automatic Lidar and Ceilometer Framework [ALCF; \citeA{kuma2021}]. We contrasted the results with the European Centre for Medium-Range Weather Forecasts (ECMWF) Reanalysis 5 [ERA5; \citeA{era5}] and the Modern-Era Retrospective analysis for Research and Applications, Version~2 [MERRA-2; \citeA{gelaro2017}].

The nextGEMS project focuses on the research and development of GSRMs at multiple modeling centers and universities in Europe. The project also develops GSRM versions of the Icosahedral Nonhydrostatic Weather and Climate Model (ICON; \citeA{hohenegger2023}), the Integrated Forecasting System [IFS; \citeA{ifs2023}], and their ocean components at eddy-resolving resolutions: ICON-O \cite{korn2022} coupled with ICON and Finite-Element/volumE Sea ice-Ocean Model [FESOM; \citeA{wang2014a}] and Nucleus for European modeling of the Ocean [NEMO; \citeA{madec2023}] coupled with IFS. The project has so far produced ICON and IFS simulations with three development versions called Cycle 1–3 and a pre-final version, with a final production version planned by the end of the project. nextGEMS is not the only project developing GSRMs; other GSRMs (or GSRM versions of climate models) currently in development include: Convection-Permitting Simulations With the E3SM Global Atmosphere Model [SCREAM; \citeA{caldwell2021}], Non-hydrostatic Icosahedral Atmospheric Model [NICAM; \citeA{satoh2008}], Unified Model (UM), eXperimental System for High-resolution modeling for Earth-to-Local Domain [X-SHiELD; \citeA{shield}], Action de Recherche Petite Echelle Grande Echelle-NonHydrostatic version [ARPEGE-NH; \citeA{bubnova1995,voldoire2017}], Finite-Volume Dynamical Core on the Cubed Sphere [FV3, \citeA{lin2004}], the National Aeronautics and Space Administration (NASA) Goddard Earth Observing System global atmospheric model version 5 [GEOS5; \citeA{putman2011}], Model for Prediction Across Scales [MPAS; \citeA{skamarock2012}], and System for Atmospheric Modeling [SAM; \citeA{khairoutdinov2003}].

Multiple cloud properties have an effect on shortwave (SW) and longwave (LW) radiation. To first order, the total cloud fraction, cloud phase, and the liquid and ice water path (LWP and IWP) are the most important cloud properties influencing SW and LW radiation. These properties are in turn influenced by the atmospheric thermodynamics, convection and circulation, and both the indirect and direct effects of aerosols. Second-order effects on SW and LW radiation are associated with the cloud droplet size distribution, ice crystal habit, cloud lifetime, and direct radiative interaction with aerosols \cite{boucher2013}. In the 6$^\mathrm{th}$ phase of the Coupled Model Intercomparison Project [CMIP6; \citeA{eyring2016}], the cloud feedback has increased relative to CMIP5 \cite{zelinka2020}, especially in the Southern Hemisphere mid-to-high latitudes, which is one of the main reasons for the higher climate sensitivity of CMIP6 models.

@@

In the following text, we refer to the SO as ocean regions south of 40°S, low-latitude SO as 40–55°S, and high-latitude SO as south of 55°S, all the way to the Antarctic coast. The reason for this dividing latitude is to split the SO into about two equal zones, as well as the results by \citeA{schuddeboom2021} (Fig.~2b) which show a contrast in CMIP model radiation biases. \citeA{schuddeboom2019} (Fig.~2) and \citeA{kuma2020} (Fig.~3) also show contrasting radiation biases in the Hadley Centre Global Environmental Model, which is also supported by \citeA{cesana2022}, displaying contrasting cloud biases due to the 0°C isotherm reaching the surface at 55°S. The findings of \citeA{niu2024}, however, support a different dividing line of 62°S based on cloud condensation nuclei concentration.

SO radiation biases have been relatively large and systematic compared to the rest of the globe since at least CMIP3 \cite{trenberth2010,bodas-salcedo2012}, and the SO SW cloud radiative effect bias is still positive in eight CMIP6 models analyzed by \citeA{schuddeboom2021} over the high-latitude SO, whereas over the low-latitude SO it tends to be more neutral or negative in some models. Too much absorbed SW radiation over the SO was also identified in the GSRM SCREAM \cite{caldwell2021}. Compensating biases are possible, such as the ``too few too bright'' cloud bias, characterized by too small a cloud fraction and too large a cloud albedo \cite{wall2017,kuma2020}, previously described by \citeA{webb2001}, \citeA{weare2004}, \citeA{zhang2005}, \citeA{karlsson2008}, \citeA{nam2012}, \citeA{klein2013}, and \citeA{bender2017} in other regions and models, which means that a model can maintain a reasonable SW radiation balance by reflecting too much SW radiation from clouds, but these cover too small an area. A study by \citeA{konsta2022} showed that this type of bias is still present in six analyzed CMIP6 models in tropical marine clouds, using the General-circulation-model-Oriented CALIPSO Cloud Product [CALIPSO–GOCCP; \citeA{chepfer2010}] and Polarization \& Anisotropy of Reflectances for Atmospheric Sciences coupled with Observations from a Lidar [PARASOL; \citeA{lier2008}] as a reference. They suggest improper simulation of subgrid-scale cloud heterogeneity as a cause. Compensating cloud biases in the Australian Community Climate and Earth System Simulator (ACCESS) – Atmosphere-only model version 2 (AM2) over the SO were analyzed by \citeA{fiddes2022} and \citeA{fiddes2024}. \citeA{possner2022} showed that over the SO, the DYAMOND GSRM ICON underestimates low-level cloud fraction on the order of 30\% and overestimates net downward TOA SW radiation by approximately 10 Wm$^\mathrm{-2}$ in the highest model resolution run (2.5~km). \citeA{zhao2022} reported a similar SW radiation bias in five analyzed CMIP6 models over the high-latitude SO and an underestimation of the total cloud fraction on the order of 10\% over the entire 40–60°S SO. Recently, \citeA{ramadoss2024} analyzed 48~hours of km-scale ICON limited-area model NWP simulations over an SO region adjacent to Tasmania against the Clouds, Aerosols, Precipitation, Radiation, and atmospherIc Composition Over the southeRn oceaN (CAPRICORN) voyage cloud and precipitation observations \cite{mcfarquhar2021}. They found the ICON cloud optical thickness was underestimated relative to Himawari‐8 satellite observations but also identified large differences in cloud top phase.

In general, sea surface temperature (SST) biases in the SO can originate either in the atmosphere \cite{hyder2018}, caused by too much SW heating of the surface or too little LW cooling of the surface, such as in situations of too much cloud cover or cloud optical thickness, or in the ocean circulation. Interactions of both are also possible; for example, SST affecting clouds and clouds affecting the surface radiation. Using ERA5 as a reference, \citeA{zhang2023} have shown that SST biases have improved in CMIP6 compared to CMIP5, with SST overall increasing in CMIP6. However, over the SO, this resulted in an even higher positive bias, especially in the Atlantic Ocean (AO) sector of the SO, increasing by up to 1°C. \citeA{luo2023} identified that the SO SST bias in an ensemble of 18 CMIP6 models originates not from the surface heat and radiation fluxes (using reanalyses as a reference) but from a warm bias in the Northern Atlantic Deep Water.

\begin{figure}[b!]
\centering
\includegraphics[width=\textwidth]{img/map_fig.pdf}
\caption{
\textbf{(a)} A map showing the tracks of 31 voyages of RV \emph{Polarstern}, RSV \emph{Aurora Australis}, RV \emph{Tangaroa}, RV \emph{Nathaniel B. Palmer}, and HMNZS \emph{Wellington} and one sub-Antarctic station (Macquarie Island) analyzed here. The tracks cover Antarctic sectors south of South America, the Atlantic Ocean, Africa, Australia, and New Zealand in the years 2010–2021 (inclusive). The dotted and dashed lines at 40°S and 55°S delineate the Southern Ocean area of our analysis and its partitioning into two subsets, respectively. A photo of \textbf{(b)} RV \emph{Polarstern} (© Folke Mehrtens, Alfred-Wegener-Institut), \textbf{(c)} Lufft CHM~15k installed on RV \emph{Tangaroa} (© Peter Kuma, University of Canterbury), \textbf{(d)} Vaisala CL51 (© Jeff Aquilina, Bureau of Meteorology), \textbf{(e)} Vaisala CT25K at Macquarie Island (© Simon P.\ Alexander, Australian Antarctic Division).
}
\label{fig:map}
\end{figure}

@@

Apart from lidar observations, radiosondes were launched on weather balloons at regular synoptic times on the RV \emph{Polarstern}, MARCUS, NBP17024, TAN1702, and TAN1802 campaigns, measuring pressure, temperature, relative humidity (RH), and the global navigation satellite system coordinates. In total, about 2300 radiosonde profiles south of 40°S were available. Spatially and temporally collocated profiles were taken from the models. Because the time period covered by the ICON model output (2021–2024) was different from the time period covered by the observations (2010–2021), when comparing with ICON, we first had to remap the observation time to model time by taking the same time relative to the start of the year. Consequently, we also had four virtual/model profiles (one for each year from 2021 to 2024) for each observed profile. Derived thermodynamic [virtual potential temperature ($\theta_v$), lifting condensation level (LCL), etc.] and dynamic physical quantities (wind speed and direction) for the measured vertical profiles were calculated with the program radiosonde tool [rstool; \citeA{rstool}]. Surface meteorological quantities were measured continuously by an onboard automatic weather station or individual instruments.

@@

An overview of the analyzed campaigns (voyages and stations). Start, end, and the number of days (UTC; inclusive) refer to the time period when the vessel was south of 40°S. Abbreviations: ceilometer (ceil.), Australia (AU), New Zealand (NZ), South America (SA), Atlantic Ocean (AO), and Africa (AF). The number of days is rounded to the nearest integer. CL51/31 indicates CL51 configured to emulate CL31. Missing days in the ceilometer data were HMNZSW16 (7 days): 24–27 November, 10~December, and 16–17~December 2016; MARCUS (3 days): 8, 10~November, and 10~December 2017; MICRE (9~days): 7–8, 29~June, 5, 16~July, 15~August, 17~October 2016, 11~February, and 21~March 2017; and TAN1502 (1~day): 24~January.

@@

A coupled (atmosphere–ocean) GSRM version of the ICON model is in development as part of the nextGEMS project \cite{hohenegger2023}. ICON is a very flexible model, allowing for simulations ranging from coarse-resolution ESM simulations, GSRM simulations, limited area model simulations, and large eddy simulations (LES) for both weather prediction and climate projections. ICON uses the atmospheric component ICON-A \cite{giorgetta2018}, whose physics is derived from ECHAM6 \cite{stevens2013}, and the ocean component ICON-O \cite{korn2022}. Earlier runs of the GSRM ICON from DYAMOND were evaluated by \citeA{mauritsen2022}.

Here, we use a free-running (i.e., the weather conditions in the model do not correspond to reality) coupled GSRM simulation made for the purpose of climate projection. nextGEMS has so far produced four cycles of model runs. We used a Cycle 3 run \emph{ngc3028} produced in 2023 \cite{nextgems2023a,nextgems2023b} for a model time period of 20 January 2020 to 22 July 2025, of which we analyzed the period 2021–2024 (inclusive). The horizontal resolution of ngc3028 is about 5 km. The model output is available on 90 vertical levels and 3-hourly instantaneous temporal resolution.

Unlike current general circulation models, the storm-resolving version of ICON does not use convective and cloud parameterization but relies on explicit simulation of convection and clouds on the model grid. Subgrid-scale clouds are not resolved, and the grid cell cloud fraction is always either 0 or 100\%. While this makes the code development simpler without having to rely on uncertain parameterizations, it can miss smaller-scale clouds below the grid resolution. Turbulence and cloud microphysics have to be parameterized in this model as in other models, and aerosols are derived from a climatology. To account for the radiative effects of subgrid-scale clouds, a cloud inhomogeneity factor is introduced in the model, which scales down the cloud liquid water for radiative calculations. It ranges from 0.4 at lower tropospheric stability (LTS) of 0~K to 0.8 at 30~K. In addition, turbulent mixing in the Smagorinsky scheme was adjusted to allow mixing or entrainment in situations of no mixing under the traditional scheme, affecting stratocumulus clouds but not trade wind clouds \cite{segura2025}.

Because the analyzed ICON simulation was free-running (years 2021–2024, inclusive), weather and climate oscillations [such as the El Niño–Southern Oscillation (ENSO) phase] are not expected to be equivalent to reality. To compare with the observations collected during a different time period (years 2010–2021, inclusive), we compared the model output with observations at the same time of year and geographical location, as determined for each data point, such as a lidar profile or a radiosonde launch. In the ALCF, this was done using the \emph{override\_year} option (\url{https://alcf.peterkuma.net/documentation/cli/cmd_model.html}). For radiosonde profiles, the same mapping of time was done. That is, when selecting an equivalent profile from the model, the time of the profile was changed so that the time relative to the start of the year was preserved, but the year was changed to one of the four years available in the model data. Thus, for every radiosonde launch, there were four equivalent model profiles. The geographical location was kept the same.
